pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-05 article.tex
Getting better

Although many possible tests can be performed, we present here only four example research questions (ERQ) that \verb;pirouette; can answer. For these ERQs we will estimate the error made by BEAST2 when using birth-death tree prior, in recovering an original tree generated under a Yule generative prior.

\begin{table}
\centering
  \begin{tabular}{|@{}c|@{}l|p{6cm}@{}|}
    \hline
    \textbf{Name} & \textbf{Meaning} & \textbf{Definition} \\ 
    \hline
    $\mathit{t}$ & tree prior & Used to simulate a DNA alignment\\
    $\mathit{c}$ & clock model & Specifies the clock for the DNA mutation rates\\
    $\mathit{s}$ & site model & Specifies the nucleotide substitution model\\
    $\mathit{\mu}$ & mutation rate & Defines the pace at which mutation occurs\\
    $\mathit{r}$ & root sequence & Specifies DNA sequence at the root of the tree\\
    $\mathit{m}$ & MRCA prior & Prior knowledge of a most recent common ancestor\\
    $\mathit{v}$ & MCMC settings & Setup of the Markov Chain Monte Carlo algorithm\\
    \hline
    $\mathit{P}$ & Generative prior & Specifies the diversification model according to which the 'true' tree is simulated\\
    $\mathit{A}$ & Alignment model & Model for the alignment simulation. It's a choice of
    $(\mathit{r},\mathit{\mu},\mathit{c},\mathit{s})$\\
    $\mathit{G}$ & Generative model & A combination of the chosen $(\mathit{P},\mathit{A})$\\
    $\mathit{I}$ & Inference model & Specifies how to execute the inference. It's a choice of
    $(\mathit{c},\mathit{s},\mathit{t},\mathit{m},\mathit{v})$\\
    $\mathit{C}$ & Inference conditions & Specifies under which conditions an inference model is used\\
    $\mathit{X}$ & Experiment & A combination of the chosen $(\mathit{I},\mathit{C})$\\
    $\mathit{E}$ & Error parameters & Sets how to measure errors\\
    $\mathit{T}$ & Twinning parameters & Specifies how to twin the 'true' tree\\
    \hline
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in this article.
  }
  \label{tab:definitions}
  \giovanni{It would be nice to specify (in the first part of the table) the basic elements composing all the elements of the second part of the table. This could be probably done for "Alignment parameters", "Inference conditions" and "Twinning parameters".}
  \giovanni{Shall we add another column with the possible option for each lower case setting? e.g. "clock model" = c("strict", "rln"); "tree prior" = c("yule", "bd")}
  \giovanni{How do we specify what constitutes 'twinning parameters'?}
\end{table}

\iffalse
Experiment & Combination of an inference model and the \\ 
    & conditions to actually use it in inference \\
    Generative model & Combination of site model, clock model and \\
    & tree prior, used to simulate a DNA alignment \\
    Inference model & Combination of site model, clock model, tree prior,\\
    & MRCA prior and MCMC, used to infer \\
    & a BEAST2 posterior \\
    MCMC & Setup of the Markov Chain Monte Carlo algorithm \\
    MRCA prior & Prior knowledge of a most recent common ancestor \\
    Site model & Model for the DNA nucleotide substitions \\
    Tree prior & Model for the macroevolutionary speciation process \\
\fi

\iffalse
$\mathit{t}$ & tree prior & Used to simulate a DNA alignment\\
    $\mathit{c}$ & clock model & Specifies the clock for the DNA \\
    & & mutation rates\\
    $\mathit{s}$ & site model & Specifies how nucleotides change\\
    & & in one another\\
    $\mathit{m}$ & MRCA prior & Prior knowledge of a most recent\\
    & & common ancestor\\
    $\mathit{v}$ & MCMC settings & Setup of the Markov Chain\\
    & &  Monte Carlo algorithm\\

    $\mathit{G}$ & Generative model & Specifies how to create the 'true'\\
    & &  tree. It's a choice of $(\mathit{c},\mathit{s},\mathit{t})$\\
    $\mathit{T}$ & Twinning parameters & Specifies how to twin the 'true' tree\\
    $\mathit{A}$ & Alignment parameters & Parameters for the alignment\\
    & & simulation\\
    $\mathit{I}$ & Inference model & Specifies how to execute the inference. \\
    & & It's a choice of
    $(\mathit{c},\mathit{s},\mathit{t},\mathit{m},\mathit{v})$\\
    $\mathit{C}$ & Inference conditions & Specifies under which conditions an\\
    & & inference model is used\\
    $\mathit{X}$ & Experiment & A combination of the chosen $(\mathit{I},\mathit{C})$\\
    $\mathit{E}$ & Error parameters & Sets how to measure errors\\
\fi

\iffalse
\richel{
  I do think this is a clumsy paragraph. I hope you can merge it in the text
}
To explain how \verb;pirouette; measures the inference error BEAST2 makes, we first give two definitions (see Table~\ref{tab:definitions} for an overview of definitions): (1) a generative model is a combination of a known tree prior (that generated a given phylogeny) and a site and clock model (that generate an alignment from the given tree), (2) an inference model is a combination of site model, clock model, tree prior, an optional most recent common-ancestor (MRCA) prior (to create a time-calibrated phylogeny) and a Markov chain Monte Carlo (MCMC) setup.
A generative model is a known/true model underlying a phylogeny and is used to simulate DNA alignments. An inference model is used to produce a posterior distribution of trees and model parameters.
\fi

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), that is converted to an alignment (2a). The user supplies one or more experiments. Every experiment is
    defined as a combination of an inference model and the conditions in which it is actually used in the inference. Based on the experiments and possibly the alignment (2a), experiments are selected to be actually used used in the next step (3a).
    Selected experiments use the alignment (2a) and their inference model 
    to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a). The posterior trees are compared to the true phylogeny (1a) using the nLTT statistic [\cite{janzen2015approximate}]. (5a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average nLTT statistic of all posterior trees in black. Taking the differences between the true tree's nLTT and each posterior's nLTT results in an error distribution, here displayed as a histogram
    of error values (6a). Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (6b) with similar intermediate steps.
  }
  \label{fig:pipeline}
    \giovanni{The black line for the average of all posterior trees is barely visible. Please, make it thicker.}
    \giovanni{Modify the figure to include elements in table~\ref{tab:definitions}. It should be specified clearly where elements corresponding to capital symbols play their roles in the pipeline. Then we should include also these elements in the caption.}
\end{figure}

\iffalse
\caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), that is converted to an alignment (2a). The user supplies one or more experiments. Every experiment is
    defined as a combination of an inference model and the conditions in which it is actually used in the inference. Based on the experiments and possibly the alignment (2a), experiments are selected to be actually used used in the next step (3a).
    Selected experiments use the alignment (2a) and their inference model 
    to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a). The posterior trees are compared to the true phylogeny (1a) using the nLTT statistic [\cite{janzen2015approximate}]. (5a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average nLTT statistic of all posterior trees in black. Taking the differences between the true tree's nLTT and each posterior's nLTT results in an error distribution, here displayed as a histogram
    of error values (6a). Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (6b) with similar intermediate steps.
  }
\fi

@@

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
This operation is performed according to the alignment model specified by the user, which consists in a root DNA sequence, a DNA mutation rate, a site model and a clock model. Since all the branches must evolve at the same pace, the clock model is always set to strict for this step. \giovanni{I think we need to specify the clock model, even though the user cannot change it. This is necessary because we refer to it when specifying what it means to set the inference model equal to the generative model. Please check if my description of why the clock can only be strict is accurate.}
This step is relatively fast, but longer DNA alignments will noticeably slow down the inference step.

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
starts from experiments as set up by the user. 
We define an experiment as the combination of an inference model 
and the conditions to actually use it in the inference step.
The inference model can be selected in two ways, either using the
generative model or selecting one or more candidates out of a list of candidate inference models.
An experiment can use the generative model as the inference model, if the generative model is known.
An experiment uses a candidate model if either the generative model is unknown,
or there is an interest in seeing how well non-generative models compete with
the generative model.
Typically, a generative model and/or a best candidate model are used in the actual inference.
When using an experiment with a generative model,
it must use the same site and clock model used in the alignment simulation,
as well as the tree prior that underlied the creation of the given phylogeny \giovanni{the tree prior appears not to be part of the alignment model. This means that should be part of the generative prior. Is it true? In case we can refer explicitly to it, according to our definitions in table \ref{tab:definitions}}. 
When using experiments with candidate models,
\verb;pirouette; selects the best candidate based
on the evidence (i.e. marginal likelihood) for the model given the alignment.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac,
therefore the selection of candidate models is not available directly on Windows. 
However, even on Windows systems there is the option to perform the model selection from browser, using \verb;mcbette;, yet the computation for this model selection is
restricted to one hour for free users of the host (Travis CI) continuous
integration service.

@@

An optional step is to generate a 'twin tree' 
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
a birth-death tree model.
\richel{Should be switched to BD, pirouette Issue 161}.
The goal of this procedure is to generate a tree
accordingly to the tree prior used in the generative model, 
that is used as a control for the main experiment.
To produce such a birth-death tree, we maximize a birth-death likelihood [\cite{nee1994reconstructed}], conditioned on having the same number of tips as the original tree, to infer its parameters.
We use these parameters, which are the per-species speciation and extinction rates, to simulate the branching times for the twin tree, which we combine with the original tree's topology.

@@

\verb;pirouette; can be installed easily from CRAN:
\rampal{be careful here; it is not on CRAN yet. Better to say it can be installed from GitHub and will be made available on CRAN)}
\richel{We adopted this policy (to say it's on CRAN in advance) with babette as well. Is this a hint to get pirouette on CRAN before the publication of this article :-) ?}
\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install.packages("pirouette")
\end{lstlisting}

@@

\richel{Suggest: use BD tree (again)} \giovanni{I think in this case we should indeed use a Yule tree prior, as we want to use a generative prior that is not birth-death (which is what BEAST2 is using for the inference). What do you think?}
We start from a tree simulated under a Yule (pure-birth) generative prior. We choose to use a tree with five taxa and a crown age of ten.

@@

The following settings are the same as in the previous section:
phylogeny (listing \ref{lst:unknown_phylogeny}), 
alignment parameters (listing \ref{lst:create_alignment}), 
experiments (listing \ref{lst:create_generative_experiment}),
BEAST2 inference and error measuring parameters \giovanni{can we add a reference to the "BEAST2 inference and error measuring parameters" we are using?}.

This time, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, 
yet its branching times are simulated according 
to the tree prior used by BEAST2 to produce a posterior. 
Since there is coherence between the inference model and the generative model, we expect BEAST2 to produce a posterior of phylogenies clearly more similar to the original one. 
We intend to use this as a control experiment.

@@

@@

\richel{check numbers}
\giovanni{I think that the current version of the discussion is too didascalic. Most of what is said can be effectively summarized by table 5. We can even add ESSes in it. Another point is that we should understand what is the message we want to deliver in this discussion. Since we are only using one tree we are not in the position to claim anything, because a single tree is not an effective sample. I saw that in some occasions you refer to the RQ as examples. This might be the right way to proceed, as our purpose is just to illustrate how to run all the possible experiments on a full tree distribution. I think we should make it clearer throughout the entire manuscript. I am also interested in hearing Rampal's opinion on this.}

@@

\giovanni{This is a first draft on how I think the discussion should be setup. In this context there is no much need for the results in the table. Let me know what you think about it.}

>>>>

Although many possible tests can be performed, 
we present here only four example research questions (ERQ) that \verb;pirouette; can answer. 
For these ERQs we will estimate the error made by BEAST2 when 
using a birth-death tree prior, in recovering an original tree 
generated under a Yule generative prior.

\begin{table}
\centering
  \begin{tabular}{|@{}c|p{3cm}|@{}l|p{6cm}@{}|}
    \hline
    \textbf{Symbol} & \textbf{Name} & \textbf{Definition} \\ 
    \hline
    $\mathit{t}$ & Tree prior & Macroevolutionary speciation model \\
    $\mathit{c}$ & Clock model & Clock for the DNA mutation rates\\
    $\mathit{s}$ & Site model & Nucleotide substitution model\\
    $\mathit{\mu}$ & Mutation rate & Pace at which mutation occurs, $\mu \in \mathit{s}$ \\
    $\mathit{r}$ & Root sequence & DNA sequence at the root of the tree\\
    $\mathit{m}$ & MRCA prior & Prior knowledge of a most recent common ancestor\\
    $\mathit{v}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup \\
    \hline
    $\mathit{P}$ & Generative prior & Diversification model the 'true' tree is simulated with \\
    $\mathit{A}$ & Alignment model & Alignment simulation model, $\mathit{A} \ni \{ \mathit{r}, \mathit{\mu}, \mathit{c}, \mathit{s} \}$, in which $\mathit{c}$ = strict \\
    $\mathit{G}$ & Generative model & Combination of a known $\mathit{P}$ and chosen (thus known) $\mathit{A}$ \\
    $\mathit{I}$ & Inference model & Bayesian phylogenetic inference model, a combination of
    $(\mathit{c},\mathit{s},\mathit{t},\mathit{m},\mathit{v})$\\
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ is used in inference \\
    $\mathit{X}$ & Experiment & Combination of $(\mathit{I},\mathit{C})$\\
    $\mathit{E}$ & Error measurement parameters & Errors measurement setup \\
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure \\
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in this article.
  }
  \label{tab:definitions}
  \richel{I've some set notations in some spots, as I am unsure if you like it}
  \giovanni{It would be nice to specify (in the first part of the table) 
    the basic elements composing all the elements of the second part of the table. 
    This could be probably done for "Alignment parameters", "Inference conditions" 
    and "Twinning parameters".
  }
  \richel{
    I feel the next table (which was also your suggestion), does so
    at the right amount of detail
  }
  \giovanni{
    How do we specify what constitutes 'twinning parameters'?
  }
  \richel{Perhaps you'd enjoy my definition?}
\end{table}

\begin{table}
\centering
  \begin{tabular}{| c | c | l |}
    \hline
    \textbf{Symbol} & \textbf{Subparameter} & \textbf{Possible values} \\ 
    \hline
    $\mathit{t}$ & name & Yule, BD, CBS, CCP, CEP \\
    $\mathit{c}$ & clock model & strict, relaxed log-normal \\
    $\mathit{s}$ & site model & JC, HKY, TN, GTR \\
    $\mathit{C}$ & model type & Generative, candidate \\
    $\mathit{C}$ & run if & Always, best candidate \\
    $\mathit{E}$ & function & nLTT, $|\gamma|$ \\
    $\mathit{T}$ & model & Yule, BD \\
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. BD = birth death, CBS = coalescent
    Bayesian Skyline, CCP = coalescent constant-population, CEP = coalescent
    exponential-population, JC = Jukes-Cantor 1969, HKY = Hasegawa, Kishino and 
    Yano 1985, TN = Tamura 1992, GTR = Generalised time-reversible 
    model (Tavaré 1986)
  }
  \label{tab:options}
\end{table}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), that is converted to an alignment (2a). The user supplies one or more experiments. Every experiment is
    defined as a combination of an inference model and the conditions in which it is actually used in the inference. Based on the experiments and possibly the alignment (2a), experiments are selected to be actually used used in the next step (3a).
    Selected experiments use the alignment (2a) and their inference model 
    to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a). The posterior trees are compared to the true phylogeny (1a) using the nLTT statistic [\cite{janzen2015approximate}]. (5a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average nLTT statistic of all posterior trees in black. Taking the differences between the true tree's nLTT and each posterior's nLTT results in an error distribution, here displayed as a histogram
    of error values (6a). Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (6b) with similar intermediate steps.
  }
  \label{fig:pipeline}
\end{figure}

@@

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
This operation is performed according to the alignment model 
specified by the user, which consists in a root DNA sequence, 
a DNA mutation rate and a site model.
A strict clock model is used as the only clock model
type, as the currently used alignment simulation algorithm
only supports a mutation rate that is equal across all the branches.
This step is relatively fast, 
but longer DNA alignments will noticeably slow down the inference step.

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
starts from experiments as set up by the user. 
We define an experiment as the combination of an inference model 
and the conditions to actually use it in the inference step.
The inference model can be selected in two ways, either using the
generative model or selecting one or more candidates out of a list of candidate inference models.
An experiment can use the generative model as the inference model, if the generative model is known.
An experiment uses a candidate model if either the generative model is unknown,
or there is an interest in seeing how well non-generative models compete with
the generative model.
Typically, a generative model and/or a best candidate model are used in the actual inference.
When using an experiment with a generative model,
it must use the same site and clock model used in the alignment simulation,
as well as the tree prior that underlied the creation of the given phylogeny 
\giovanni{
  the tree prior appears not to be part of the alignment model. 
  This means that should be part of the generative prior. 
  Is it true? In case we can refer explicitly to it, 
  according to our definitions in table \ref{tab:definitions}
}
\richel{
  Indeed, a tree prior is not part of the alignment model.
  Go ahead and show how you'd like to refer explicitly to the
  table!
}
. 
When using experiments with candidate models,
\verb;pirouette; selects the best candidate based
on the evidence (i.e. marginal likelihood) for the model given the alignment.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac,
therefore the selection of candidate models is not available directly on Windows. 
However, even on Windows systems there is the option to perform the model selection from browser, using \verb;mcbette;, yet the computation for this model selection is
restricted to one hour for free users of the host (Travis CI) continuous
integration service.

@@

An optional step is to generate a 'twin tree' 
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
a birth-death tree model.
The goal of this procedure is to generate a tree
accordingly to the tree prior used in the generative model, 
that is used as a control for the main experiment.
To produce such a birth-death tree, we maximize a birth-death likelihood [\cite{nee1994reconstructed}], conditioned on having the same number of tips as the original tree, to infer its parameters.
We use these parameters, which are the per-species speciation and extinction rates, to simulate the branching times for the twin tree, which we combine with the original tree's topology.

@@

\verb;pirouette; can be installed easily from CRAN:
\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install.packages("pirouette")
\end{lstlisting}

@@

\giovanni{
  I think in this case we should indeed use a Yule tree prior [...]. What do you think?}
\richel{
  Agrees. I was wrong, possibly due to carnaval. Yule is the way to go!
}
We start from a tree simulated under a Yule (pure-birth) generative prior. 
We choose to use a tree with five taxa and a crown age of ten.

@@

Most of the settings for this example research question are
the same as in the previous section, which are
the phylogeny (listing \ref{lst:unknown_phylogeny}), 
the alignment parameters (listing \ref{lst:create_alignment}) 
and the list of experiments (listing \ref{lst:create_generative_experiment}).

This time, however, we are interested in creating a twin tree,
as such a twin tree allows us to measure a baseline error produced
by the BEAST2 run.
A twin tree
has the same topology as the given tree, 
yet its branching times are simulated according 
to a tree prior already implemented by BEAST2 to produce a posterior. 
Because for a twin tree we know the tree prior 
and because we use that tree prior in inference, 
we expect BEAST2 to produce a posterior of phylogenies
that is close to the twin tree.
As the inference model for the twin tree is (more) completely
specified than the true tree, 
we expect that a twin tree and its posterior trees are more similar
than a true tree and its posterior trees.
Due to this, the twinning is used a control for the experiment.

@@

\giovanni{describe winner}
\richel{indeed, after the figures also show it :-) }

@@

\richel{check numbers}
\giovanni{
  I think that the current version of the discussion is too didascalic. 
  Most of what is said can be effectively summarized by table 5. 
  We can even add ESSes in it. 
}
\giovanni{
  Another point is that we should understand what is 
  the message we want to deliver in this discussion. 
  Since we are only using one tree we are not in the position to claim anything, 
  because a single tree is not an effective sample. 
  I saw that in some occasions you refer to the RQ as examples. 
  This might be the right way to proceed, 
  as our purpose is just to illustrate how to run all the possible experiments 
  on a full tree distribution. 
  I think we should make it clearer throughout the entire manuscript. 
  I am also interested in hearing Rampal's opinion on this.
}

@@

\giovanni{
  This is a first draft on how I think the discussion should be setup. 
  In this context there is no much need for the results in the table. 
  Let me know what you think about it.
}
